Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-05-06 paper/ifo-appendix.tex
another set of edits on all three documents

The intensity $I(t)$ at $P$ is then the norm squared of the total electric field amplitude,
\begin{equation}
I(t) = \lvert A e^{i (k L - \omega t)}  + A e^{i (k (L + 2 d \cos{\theta}) - \omega t)} \rvert^2\,.
\label{eqn:intensity1}    
\end{equation}
Some elementary trigonometry yields
\begin{equation}
%I(t) = 2 A^2 (1 + \cos{\phi})
I(t) = 2 A^2 \left[ 1 + \cos{\left( 2 k d(t) \cos{\theta} \right) } \right]  \label{eqn:intensity2}\,.
\end{equation}
Equation~\ref{eqn:intensity2} shows that $I(t)$ and $d(t)$ are related nonlinearly. 
Near a bright fringe, where one has $\cos{[2kd\cos{(\theta)}]} \approx 1$, the nonlinearity is quadratic, with $\delta I \propto (\delta d)^2$ for small $\delta d$. 
Consequently, an oscillation of the form $\delta d \propto \exp{( -i \Omega t )}$ in the arm length difference produces a DC signal and a frequency-doubled signal $\propto \exp{(-2i\Omega t)}$ in $\delta I$. 
This is discussed further in Section~\ref{sec:ifo}.

@@

%\subsection{Pseudocode}
Here we present some pseudocode of the implementation used in Section~\ref{sec:viterbi}. More general pseudocode complete with technical language is available elsewhere online~\cite{viterbipseudocode:online}. We use Fourier amplitudes, normalized between $(0, 1)$ by dividing by the maximum value in the grid, as multiplicative weights. To avoid numerical underflow we take the logarithm and consider them instead as additive weights.

Let $X$ be grid of $j=0,\ldots,N_f$ rows and $i=0,\ldots,N_t$ columns of additive weights for each node. Let Y and Z be grids of the same shape to store the weight of the best path to each node and the row index of the previous node on that path, respectively. Let W be a length $N_t+1$ array to store the final sequence of row indices for the optimal path. We restrict paths to only move one cell up or down at a time (or stay constant). For the edge cases of $j=0,N_f$, we only search over $k \in \{0,1\}$ and $k \in \{-1,0\}$, respectively, to stay inside the grid (this is not shown in the pseudocode).

@@

\begin{figure*}
	\includegraphics[width=.8\textwidth]{figures/circuit_diagram_2.pdf}
	\caption{\label{fig:circuit_diagram}
Circuit diagram for reading the photodiode. 
The photodiode is connected in reverse-bias across an op-amp. 
The op-amp output is passed through a Sallen-Key anti-aliasing filter tuned to $16\,{\rm kHz}$, then into an analog-to-digital converter (ADC). 
The digitized signal is then read by the special purpose input (SPI) pins of a Raspberry Pi in standard configuration.
This diagram is produced using the Circuit Diagram online tool~\cite{CircuitDiagram:online} and is available from~\url{https://crcit.net/c/e397dcc2166943d69155f9dac1e27bce}.
}
\end{figure*}

>>>>

The intensity $I(t)$ at $P$ is then the norm squared of the total electric field amplitude,
\begin{equation}
I(t) = \left| A e^{i (k L - \omega t)}  + A e^{i (k (L + 2 d \cos{\theta}) - \omega t)} \right|^2\,.
\label{eqn:intensity1}    
\end{equation}
Some elementary trigonometry yields
\begin{equation}
%I(t) = 2 A^2 (1 + \cos{\phi})
I(t) = 2 A^2 \left[ 1 + \cos{\left( 2 k d(t) \cos{\theta} \right) } \right]  \label{eqn:intensity2}\,.
\end{equation}
Equation~\ref{eqn:intensity2} shows that $I(t)$ and $d(t)$ are related nonlinearly. 
Near a bright fringe, where one has $\cos{[2kd\cos{(\theta)}]} \approx 1$, the nonlinearity is quadratic, with $\delta I \propto (\delta d)^2$ for small $\delta d$. 
Consequently, an oscillation of the form $\delta d \propto \exp{( -i \Omega t )}$ in the arm length difference produces a DC signal and a frequency-doubled signal $\propto \exp{(-2i\Omega t)}$ in $\delta I$. 
This is discussed further in Section~\ref{sec:ifo}.

@@

%\subsection{Pseudocode}
Here, we present some pseudocode of the implementation used in Section~\ref{sec:viterbi}. More general pseudocode complete with technical language is available elsewhere online~\cite{viterbipseudocode:online}. We use Fourier amplitudes, normalized between $(0, 1)$ by dividing by the maximum value in the grid, as multiplicative weights. To avoid numerical underflow we take the logarithm and consider them instead as additive weights.

Let $X$ be grid of $j=0,\ldots,N_f$ rows and $i=0,\ldots,N_t$ columns of additive weights for each node. Let Y and Z be grids of the same shape to store the weight of the best path to each node and the row index of the previous node on that path, respectively. Let W be a length $N_t+1$ array to store the final sequence of row indices for the optimal path. We restrict paths to only move one cell up or down at a time (or stay constant). For the edge cases of $j=0,N_f$, we only search over $k \in \{0,1\}$ and $k \in \{-1,0\}$, respectively, to stay inside the grid (this is not shown in the pseudocode below).

@@

\begin{figure*}
	\includegraphics[width=.8\textwidth]{figures/circuit_diagram_2.pdf}
	\caption{\label{fig:circuit_diagram}
Circuit diagram for reading the photodiode. 
The photodiode is connected in reverse-bias across an op-amp (pale-red box). 
The op-amp output is passed through a Sallen-Key anti-aliasing filter (pale-blue box) tuned to $16\,{\rm kHz}$, then into an analog-to-digital converter (ADC). 
The digitized signal is then read by the special-purpose input (SPI) pins of a Raspberry Pi in standard configuration.
This diagram is produced using the Circuit Diagram online tool~\cite{CircuitDiagram:online} and is available from~\url{https://crcit.net/c/e397dcc2166943d69155f9dac1e27bce}.
}
\end{figure*}

### 2021-05-06 paper/ifo-complexAudio.tex
responding to Andrew's comments on Supp. Material
another set of edits on all three documents

This analysis is indirectly related to the advanced signal processing techniques required to extract gravitational-wave signals. 
In addition, the apparatus serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories. 
We describe additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and the initial results in Section~\ref{sec:initialResultsOpMic}. 
We consider a selection of filter techniques, details of which can be found in the Supplementary Material and we refer the reader to Refs.~\cite{Mitra:2011, Lyons:2011, Stein:2000, OpenheimSchaferBuck:1999, PrakisManolakis:1996,10.5555/151045} for further information on digital signal processing. 
In Section~\ref{sec:opticalMicResults} we present the two best-performing analyses from our Supplementary Material.

\subsection{Hardware modifications for the optical microphone}
\label{sec:photodiode}
The human ear can hear frequencies in the range of $\sim 20\,{\rm Hz}$--$20\,{\rm kHz}$. 
Speech intelligibility (being able to understand speech) requires frequencies up to $3\,{\rm kHz}$ and music requires up to and beyond $8\,{\rm kHz}$~\cite{speech_intelligibility}. Therefore, the optical microphone requires a sample rate of at least $16\,{\rm kHz}$ to capture both speech and music (adjusting for the Nyquist frequency). 
This cannot be achieved with the webcam used in Sections~\ref{sec:single_tone} and ~\ref{sec:viterbi_wandering} as it has a sampling rate of $30\,{\rm Hz}$ and thus can only ``hear'' frequencies below $15\,{\rm Hz}$.
To overcome this issue, we use a photodiode\footnote{A photodiode is an electrical component that acts as a regular diode when no light is incident on it, blocking any current flow in the reverse direction. As the intensity of incident light rises, it becomes increasingly conductive in the reverse direction.} at the output of the interferometer to achieve a sampling rate of $16\,{\rm kHz}$.

@@

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing\footnote{Folding of frequencies greater than half the sampling rate.} into the detected range. We include an anti-aliasing Sallen-Key filter~\cite{sallen_key_filter} tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$, before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was re-purposed grill cloth from a commercial speaker.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

Environmental noise reduction is an active area of research for gravitational-wave detectors~\cite{EfflerEtAl:2015}.
In some cases environmental noise sources can be removed or mitigated on site~\cite{CovasEtAl:2018}. 
Where this is not possible, noise subtraction can be done after data collection~\cite{DriggersEtAl:2019, DavisEtAl:2019}. 
One example is removal of spectral noise lines from the United States power grid operating at $60\,{\rm Hz}$~\cite{VajenteEtAl:2020}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.95\textwidth]{figures/combined_highlight_results_melatos_labelled.pdf}
\caption{\label{fig:notchWienerLogMMSEResults}
Timeseries (left column) and frequency spectrum (right column) results with the optical microphone. 
The original input signal in the first row is a $1\,{\rm s}$ recording of an adult male voice (saying ``A cathode''). 
 The input signal is shifted by 0.12 s to the right to synchronize the manual delay from starting the recording with the Raspberry Pi and starting to play the source through the speaker. 
The second row shows the raw output from the optical microphone when the input from the first row is played. 
The third row shows the result of applying the notch and Wiener filter combined. 
The fourth row shows the result of applying the logMMSE estimator, where the rise at the start of the timeseries is an expected effect when filtering a signal of finite duration. 
}
\end{center}
\end{figure*}

@@

In this section, we present results of two advanced signal processing analysis techniques applied to the optical microphone recordings. 
The techniques are only briefly described here and we refer the reader to the accompanying Supplementary Material for further details and some other analysis techniques.

@@

Secondly, we apply a speech enhancement technique. 
Ref.~\cite{SubjectiveComparison} compares $13$ speech enhancement methods, finding the log minimum mean-square-error (logMMSE) estimator to be the best, qualitatively, at recovering speech (see also Ref.~\cite{Ephraim1984SpeechEU_logMMSE} for details on the logMMSE estimator). 
We use an existing implementation of the logMMSE from~\cite{logmmse}.  
The logMMSE estimator results are shown in the bottom panel of Fig.~\ref{fig:notchWienerLogMMSEResults}. 
We see significant attenuation of the mains harmonics and general smoothing of the spectrum. 
Most of the background noise is removed, however, the logMMSE does not significantly enhance the speech as the voice sounds muffled and indistinct.

>>>>

This analysis is indirectly related to the advanced signal processing techniques required to extract gravitational-wave signals. 
In addition, the apparatus serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories. 
We describe additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and the initial results in Section~\ref{sec:initialResultsOpMic}. 
We consider a selection of filter techniques, details of which can be found in the Supplementary Material and we refer the reader to Refs.~\cite{Mitra:2011, Lyons:2011, Stein:2000, OpenheimSchaferBuck:1999, PrakisManolakis:1996,10.5555/151045} for further information on digital signal processing. 
In Section~\ref{sec:opticalMicResults}, we present the two best-performing analyses from our Supplementary Material.

\subsection{Hardware modifications for the optical microphone}
\label{sec:photodiode}
The human ear can hear frequencies in the range of $\sim 20\,{\rm Hz}$--$20\,{\rm kHz}$. 
Speech intelligibility (being able to understand speech) requires frequencies up to $3\,{\rm kHz}$ and music requires up to and beyond $8\,{\rm kHz}$~\cite{speech_intelligibility}. Therefore, the optical microphone requires a sample rate of at least $16\,{\rm kHz}$ to capture both speech and music (adjusting for the Nyquist frequency). 
This cannot be achieved with the webcam used in Sections~\ref{sec:single_tone} and ~\ref{sec:viterbi_wandering} as it has a sampling rate of $30\,{\rm Hz}$ and thus can only ``hear'' frequencies below $15\,{\rm Hz}$.
To overcome this issue, we use a photodiode~\footnote{A photodiode is an electrical component that acts as a regular diode when no light is incident on it, blocking any current flow in the reverse direction. As the intensity of incident light rises, it becomes increasingly conductive in the reverse direction.} at the output of the interferometer to achieve a sampling rate of $16\,{\rm kHz}$.

@@

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing~\footnote{Folding of frequencies greater than half the sampling rate.} into the detected range. We include an anti-aliasing Sallen-Key filter~\cite{sallen_key_filter} tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$ before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was re-purposed grill cloth from a commercial speaker.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

@@

Environmental noise reduction is an active area of research for gravitational-wave detectors~\cite{EfflerEtAl:2015}.
In some cases, environmental noise sources can be removed or mitigated on site~\cite{CovasEtAl:2018}. 
Where this is not possible, noise subtraction can be done after data collection~\cite{DriggersEtAl:2019, DavisEtAl:2019}. 
One example is the removal of spectral noise lines from the United States power grid operating at $60\,{\rm Hz}$~\cite{VajenteEtAl:2020}. %United States's?

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/combined_highlight_results_melatos_labelled.pdf}
\caption{\label{fig:notchWienerLogMMSEResults}
Timeseries (left column) and frequency spectrum (right column) results with the optical microphone. 
The original input signal in the first row is a $1\,{\rm s}$ recording of an adult male voice (saying ``A cathode''). 
 The input signal is shifted by 0.12 s to the right to synchronize the manual delay from starting the recording with the Raspberry Pi and starting to play the source through the speaker. 
The second row shows the raw output from the optical microphone when the input from the first row is played. 
The third row shows the result of applying the notch and Wiener filter combined. 
The fourth row shows the result of applying the logMMSE estimator, where the rise at the start of the timeseries is an expected effect when filtering a signal of finite duration. 
}
\end{center}
\end{figure*}

@@

In this section, we present the results of two advanced signal processing analysis techniques applied to the optical microphone recordings. 
The techniques are only briefly described here and we refer the reader to the accompanying Supplementary Material for further details and some other analysis techniques.

@@

Secondly, we apply a speech enhancement technique. 
Ref.~\cite{SubjectiveComparison} compares $13$ speech enhancement methods, finding the log minimum mean-square error (logMMSE) estimator to be the best, qualitatively, at recovering speech (see also Ref.~\cite{Ephraim1984SpeechEU_logMMSE} for details on the logMMSE estimator). 
We use an existing implementation of the logMMSE from~\cite{logmmse}.  
The logMMSE estimator results are shown in the bottom panel of Fig.~\ref{fig:notchWienerLogMMSEResults}. 
We see significant attenuation of the mains harmonics and general smoothing of the spectrum. 
Most of the background noise is removed, however, the logMMSE does not significantly enhance the speech as the voice sounds muffled and indistinct.

### 2021-05-06 paper/ifo-conclusions.tex
another set of edits on all three documents

In this paper, we use a table-top Michelson interferometer as an analog to a gravitational-wave detector, demonstrating signal processing techniques used within the gravitational-wave community and more broadly. 
Gravitational-wave signals are simulated by playing sound recordings through a speaker fixed to the back of one of the arm mirrors. 
We consider quasi-monochromatic audio signals as an analog to continuous-wave signals to directly demonstrate techniques used in current continuous gravitational-wave searches.
We then consider more complex audio (music and speech) for a more general interest in signal processing and explore the use of the interferometer as an optical microphone. 
A variety of example activities are considered in the Supplementary Material. 
We find that the logMMSE produces the best results, however, it does not enhance the recording to an intelligible level. 
The complex audio signals can also serve as a more distant analog to un-modeled gravitational-wave burst signals, e.g. from supernovae~\cite{MillhouseEtAl:2018,SuvorovaPowellMelatos:2019}.

@@

The demonstration presented has the potential to be adapted for use in the undergraduate laboratory.
It can be used to teach topics of interest to physics and electrical engineering students including, interferometer physics; gravitational-wave detection, searches, and analysis; and signal processing with filters and speech enhancement techniques. 
The increased excitement and public interest in the field of gravitational-wave research in recent years mean that this demonstration may also be adapted as a tool for explaining gravitational-wave research to a wider, non-specialist audience. 
The field of gravitational-wave astrophysics continues to grow and the future will bring many more detections of binary black holes and neutron stars, as well as the anticipated first detection of other classes of signals, such as continuous waves, to which this demonstration provides some charming insights. 
% \jam{``charming'' is very sweet}     - agreed I wasn't sure at first but it has grown on me a lot! :) - Hannah
%The demonstration presented here has great potential to be adapted for use in the undergraduate laboratory. It can be used to demonstrate topics such as interferometer physics; gravitational wave detection, searches, and analysis; and signal processing with filters and speech enhancement techniques to physics and electrical engineering students. Given the increased level of excitement and public interest in the field of gravitational wave astronomy in recent years, this table-top demonstration can be used as a tool for explaining gravitational wave research to a wider, non-specialist audience. The future will bring many more detections, including further observations of transient signals from binary black hole and binary neutron star mergers, as well as the anticipated first detections of other classes of signal, including continuous waves, to which this demonstration provides some charming insights.

>>>>

In this paper, we use a table-top Michelson interferometer as an analog to a gravitational-wave detector, demonstrating signal processing techniques used within the gravitational-wave community and more broadly. 
Gravitational-wave signals are simulated by playing sound recordings through a speaker fixed to the back of one of the arm mirrors. 
We consider quasi-monochromatic audio signals as an analog to continuous-wave signals to directly demonstrate techniques used in current continuous gravitational-wave searches.
We then consider more complex audio (music and speech) for a more general interest in signal processing and explore the use of the interferometer as an optical microphone. 
A variety of example activities is considered in the Supplementary Material. 
We find that the logMMSE produces the best results, however, it does not enhance the recording to an intelligible level. 
The complex audio signals can also serve as a more distant analog to un-modeled gravitational-wave burst signals, e.g. from supernovae~\cite{MillhouseEtAl:2018,SuvorovaPowellMelatos:2019}.

@@

The demonstration presented has the potential to be adapted for use in the undergraduate laboratory.
It can be used to teach topics of interest to physics and electrical engineering students including interferometer physics; gravitational-wave detection, searches, and analysis; and signal processing with filters and speech enhancement techniques. 
The increased excitement and public interest in the field of gravitational-wave research in recent years mean that this demonstration may also be adapted as a tool for explaining gravitational-wave research to a wider, non-specialist audience. 
The field of gravitational-wave astrophysics continues to grow and the future will bring many more detections of binary black holes and neutron stars, as well as the anticipated first detection of other classes of signals, such as continuous waves, to which this demonstration provides some charming insights. 
% \jam{``charming'' is very sweet}     - agreed I wasn't sure at first but it has grown on me a lot! :) - Hannah
%The demonstration presented here has great potential to be adapted for use in the undergraduate laboratory. It can be used to demonstrate topics such as interferometer physics; gravitational wave detection, searches, and analysis; and signal processing with filters and speech enhancement techniques to physics and electrical engineering students. Given the increased level of excitement and public interest in the field of gravitational wave astronomy in recent years, this table-top demonstration can be used as a tool for explaining gravitational wave research to a wider, non-specialist audience. The future will bring many more detections, including further observations of transient signals from binary black hole and binary neutron star mergers, as well as the anticipated first detections of other classes of signal, including continuous waves, to which this demonstration provides some charming insights.
